Sleep, relative to wake, increases both veridical and false memory in the DRM paradigm: A registered report

### 2023-07-09 manuscript.Rmd
Added code for semantic distance analysis

packagelist <- c('bookdown','knitr','markdown','osfr','plyr','tidyr','dplyr','emmeans','ggplot2','Rmisc','buildmer','lme4','lmerTest','stringr','effects','effectsize','ggdist','car','gridExtra','kableExtra','rstatix')
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

ddply(lure_pp_explore_delay, .(Test.Time,adjusted_recall_binary), summarise, average=mean(Lure.Recalled), stdev=sd(Lure.Recalled))

@@

if (processdata==1){
iota_buildmer000 <- buildmer(lure_recalled ~ Delay*Test.Time*N_per_list+ intrusions +
                               (N_per_list+ intrusions | Participant.Private.ID) + 
                               (Test.Time| lure), 
                           data = iota_explore, 
                           family = "binomial",
                           buildmerControl = buildmerControl(direction='backward',args =list(control=glmerControl(optimizer="bobyqa"))))

@@

```{r delaygroup}
if (processdata){
iota_buildmer2 <- buildmer(lure_recalled ~ Test.Time*N_per_list+ intrusions +
                                       (N_per_list+ intrusions | Participant.Private.ID) + 
                                       (Test.Time| lure), 
                                     data = iota_explore_delay, 
                                     family = "binomial",
                                     buildmerControl = buildmerControl(direction='backward',args =list(control=glmerControl(optimizer="bobyqa"))))

@@

if (processdata){
iota_explore_immediate <- subset(iota_explore, Delay=="Immediate/Control")

iota_buildmer3_standard <- glmer(lure_recalled ~Test.Time*N_per_list+intrusions+
                                       (1 | Participant.Private.ID) + 
                                       (1| lure), data = iota_explore_immediate, family = binomial(link = 'logit'), control = glmerControl(optimizer = 'bobyqa'))

@@

@@

@@

### Complementary Bayesian analysis
Although our inference was based on a frequentist approach, we pre-registered to use a Bayesian analysis to complement and test the strength of our results.

@@

In this exploratory analysis, we followed Diekelmann et al. [-@diekelmann2010a] by focusing on the Delay groups\footnote{Note that Diekelmann et al. (2010) did not have Immediate/Control groups.} and by performing a median split on our sample’s adjusted veridical recall (Median = 5), classifying our participants as either high (>5; N = 117) or low ( $\leq$ 5; N = 123) performers. We followed the analysis approach of Diekelmann et al. [-@diekelmann2010a], but since the number of critical lures a participant produced is count data, we used a 2 (Group: Sleep vs. Wake) $\times$ 2 (Adjusted veridical recall: High vs. Low Performers) Poisson regression, instead of an ANOVA. This analysis showed no main effect of Group ($\beta$ = -0.052, z = -0.494, p = .621) but a main effect of Adjusted veridical recall ($\beta$ = -0.438, z = -3.766, p < .001), such that high performers (M = 3.17, SD = 2.71) tended to falsely recall more critical lures than low performers (M = 2.09, SD = 1.68) ([@thapar2001a; @toglia1999], but see [@cann2011a; @roediger2001a; @stadler1999a]). Furthermore, importantly, the interaction between Group and Adjusted veridical recall was not significant ($\beta$ = 0.049, z = 0.30, p = .763). Despite this, we followed Diekelmann et al. [-@diekelmann2010a] by comparing low performers in the Sleep and Wake groups. Contrary to their findings, emmeans showed that our low performers in both groups recalled essentially the same number of critical lures (MSleep = 2.09, SDSleep = 1.63 vs. Mwake = 2.09, SDWake = 1.74; $\beta$ = 0.133, z = 0.021, p = .982). In other words, this exploratory analysis found no evidence that the effect of sleep on false recall was modulated by veridical memory, casting doubt over the robustness and reliability of Diekelmann et al.’s [-@diekelmann2010a] findings.

@@

Across the Immediate and Delay groups, the number of intrusions and correct recall per list both had a main effect (ps $\leq$ .001) such that they were positively correlated with lure recall. However, the effect of Test Time was not significant in either group (ps > .11). Finally, for the Correct recall per list $\times$ Test Time interaction, it was significant in the Delay (z = -3.504, p < .001) but not the Immediate groups (z = -0.037, p = .971). To interpret the former, we used the R package, effects [@fox2022a], to visualise it (see Figure \@ref(fig:delayfig)A) and plotted lure recall probability against veridical recall on a participant level (see Figure \@ref(fig:delayfig)B).

>>>>

packagelist <- c('bookdown','knitr','markdown','osfr','plyr','tidyr','dplyr','emmeans','ggplot2','Rmisc','buildmer','lme4','lmerTest','stringr','effects','effectsize','ggdist','car','gridExtra','kableExtra','rstatix','LSAfun')
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

premmeans <- ddply(lure_pp_explore_delay, .(Test.Time,adjusted_recall_binary), summarise, average=mean(Lure.Recalled), stdev=sd(Lure.Recalled))

@@

iota_buildmer000 <- buildmer(lure_recalled ~ Delay*Test.Time*N_per_list+ intrusions + (N_per_list+ intrusions | Participant.Private.ID) + (Test.Time| lure), data = iota_explore, family = "binomial", buildmerControl = buildmerControl(direction='backward', args = list(control=glmerControl(optimizer="bobyqa"))))

@@

```{r delaygroup}

if (processdata){

iota_buildmer2 <- buildmer(lure_recalled ~ Test.Time*N_per_list+ intrusions +
(N_per_list+ intrusions | Participant.Private.ID) + (Test.Time| lure), data = iota_explore_delay, family = "binomial",buildmerControl = buildmerControl(direction='backward',args = list(control=glmerControl(optimizer="bobyqa"))))

@@

if (processdata){

iota_explore_immediate <- subset(iota_explore, Delay=="Immediate/Control")

iota_buildmer3_standard <- glmer(lure_recalled ~ Test.Time * N_per_list + intrusions + (1 | Participant.Private.ID) + (1| lure), data = iota_explore_immediate, family = binomial(link = 'logit'), control = glmerControl(optimizer = 'bobyqa'))

@@

```{r semanticdistance}

# processdata <- 1

osfproject <- osf_retrieve_node("9pdyf")
osffiles <- osf_ls_files(osfproject,path='Stage 2/Data/Semantic distance (exploratory analysis)/')
if (!file.exists('data/ukwac_cbow.rda')){
  hid <- pmatch('ukwac_cbow.rda',as.character(unlist(osffiles[,1])))
  osf_download(osffiles[hid,], path='data/')
}

load('data/ukwac_cbow.rda')

lure_item <- head(fm$word, 20)

intrusion_only <- subset(fm, intrusion==1)
intrusion_item <- intrusion_only$response

my_list <- vector()
for (i in 1:length(intrusion_item)){
  new_element <- Cosine(lure_item[1], intrusion_item[i], ukwac)
  my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[2], intrusion_item[i], ukwac)
  my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[3], intrusion_item[i], ukwac)
  my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[4], intrusion_item[i], ukwac)
  my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[5], intrusion_item[i], ukwac)
  my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[6], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[7], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[8], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[9], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[10], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[11], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[12], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[13], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[14], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[15], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[16], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[17], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[18], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[19], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  new_element <- Cosine(lure_item[20], intrusion_item[i], ukwac)
   my_list[[length(my_list) + 1]] <- new_element
  }

my_list <- as.data.frame(my_list)
my_list$item <- rep(1:3301, each=20)
closest <- ddply(my_list, .(item), summarise, closest=max(my_list))

average_closest <- ddply(intrusion_with_distance, .(Participant.Private.ID, Delay, Test.Time), summarise, meanDistance=mean(closest))

ddply(average_closest, .(Delay, Test.Time), summarise, average=mean(meanDistance, na.rm=T), stdev=sd(meanDistance, na.rm = T))

semsum <- summary(semantic_ANOVA <- aov(meanDistance~Delay*Test.Time, data=average_closest))
semmeans <- emmeans(semantic_ANOVA, pairwise ~ Test.Time | Delay)

save(file='data/processeddata.RData',list=c('intrusion_poisson','intrusionout','intrusion_CI','lure_model_buildmer','lure_model','lure_pairwise','lureeffects','lure_model_without_covariate_buildmer','null_lure','lureBF','lure_delay_model1','lure_immediate_model1','lure_delay_model1_null','lure_immediate_model1_null','luredelayBF','lureimmediateBF','lurettest','studied_model_buildmer','studied_model','studiedem1','studiedem2','studiedBF','studied_delay_model1','studied_delay_model1_null','studdelayBF','studied_immediate_model1','studied_immediate_model1_null','studimBF','prem','prsum','iota_buildmer000','iota_buildmer2','iota_buildmer2_standard','iota_buildmer3_standard','my_list','closest','average_closest','semsum','semmeans'))

@@

temp <- capture.output(prem$contrasts[1])
emout5 <- unlist(strsplit(gsub("([*]+$)|[<]", "", temp[2]), "\\s+"))

@@

### Complementary Bayesian analysis

Although our inference was based on a frequentist approach, we pre-registered to use a Bayesian analysis to complement and test the strength of our results.

@@

In this exploratory analysis, we followed Diekelmann et al. [-@diekelmann2010a] by focusing on the Delay groups\footnote{Note that Diekelmann et al. (2010) did not have Immediate/Control groups.} and by performing a median split on our sample’s adjusted veridical recall (Median = 5), classifying our participants as either high (>5; N = 117) or low ($\leq$ 5; N = 123) performers. We followed the analysis approach of Diekelmann et al. [-@diekelmann2010a], but since the number of critical lures a participant produced is count data, we used a 2 (Group: Sleep vs. Wake) $\times$ 2 (Adjusted veridical recall: High vs. Low Performers) Poisson regression, instead of an ANOVA. This analysis showed no main effect of Group ($\beta$ = `r round(as.numeric(emout5[7]),digits=3)`, z = `r round(as.numeric(emout5[10]),digits=3)`, p = `r p_format(as.numeric(emout5[11]),digits=3,leading.zero=F)`) but a main effect of Adjusted veridical recall ($\beta$ = -0.438, z = -3.766, p < .001), such that high performers (M = 3.17, SD = 2.71) tended to falsely recall more critical lures than low performers (M = 2.09, SD = 1.68) ([@thapar2001a; @toglia1999], but see [@cann2011a; @roediger2001a; @stadler1999a]). Furthermore, importantly, the interaction between Group and Adjusted veridical recall was not significant ($\beta$ = 0.049, z = 0.30, p = .763). Despite this, we followed Diekelmann et al. [-@diekelmann2010a] by comparing low performers in the Sleep and Wake groups. Contrary to their findings, emmeans showed that our low performers in both groups recalled essentially the same number of critical lures (M$_{Sleep}$ = `r round(premmeans[2,3],digits=2)`, SD$_{Sleep}$ = `r round(premmeans[2,4],digits=2)` vs. M$_{Wake}$ = `r round(premmeans[4,3],digits=2)`, SD$_{Wake}$ = `r round(premmeans[4,4],digits=2)`; $\beta$ = 0.133, z = 0.021, p = .982). In other words, this exploratory analysis found no evidence that the effect of sleep on false recall was modulated by veridical memory, casting doubt over the robustness and reliability of Diekelmann et al.’s [-@diekelmann2010a] findings.

@@

Across the Immediate and Delay groups, the number of intrusions and correct recall per list both had a main effect (ps $\leq$ .001) such that they were positively correlated with lure recall. However, the effect of Test Time was not significant in either group (ps > .11). Finally, for the Correct recall per list $\times$ Test Time interaction, it was significant in the Delay (z = `r round(modout2[5,3],digits=3)`, p `r p_format(modout2[5,4],digits=3,accuracy=0.001,leading.zero=F)`) but not the Immediate groups (z = `r round(modout1[5,3],digits=3)`, p = `r p_format(modout1[5,4],digits=3,accuracy=0.001,leading.zero=F)`). To interpret the former, we used the R package, effects [@fox2022a], to visualise it (see Figure \@ref(fig:delayfig)A) and plotted lure recall probability against veridical recall on a participant level (see Figure \@ref(fig:delayfig)B).

### 2023-07-09 manuscript.tex
Added code for semantic distance analysis

In this exploratory analysis, we followed Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{34}{]} by focusing on the Delay groups\footnote{Note that Diekelmann et al. (2010) did not have Immediate/Control groups.} and by performing a median split on our sample's adjusted veridical recall (Median = 5), classifying our participants as either high (\textgreater5; N = 117) or low ( \(\leq\) 5; N = 123) performers. We followed the analysis approach of Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{34}{]}, but since the number of critical lures a participant produced is count data, we used a 2 (Group: Sleep vs.~Wake) \(\times\) 2 (Adjusted veridical recall: High vs.~Low Performers) Poisson regression, instead of an ANOVA. This analysis showed no main effect of Group (\(\beta\) = -0.052, z = -0.494, p = .621) but a main effect of Adjusted veridical recall (\(\beta\) = -0.438, z = -3.766, p \textless{} .001), such that high performers (M = 3.17, SD = 2.71) tended to falsely recall more critical lures than low performers (M = 2.09, SD = 1.68) ({[}\protect\hyperlink{ref-thapar2001a}{\textbf{thapar2001a?}},\protect\hyperlink{ref-toglia1999}{\textbf{toglia1999?}}{]}, but see {[}\protect\hyperlink{ref-roediger2001a}{21},\protect\hyperlink{ref-stadler1999a}{35},\protect\hyperlink{ref-cann2011a}{\textbf{cann2011a?}}{]}). Furthermore, importantly, the interaction between Group and Adjusted veridical recall was not significant (\(\beta\) = 0.049, z = 0.30, p = .763). Despite this, we followed Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{34}{]} by comparing low performers in the Sleep and Wake groups. Contrary to their findings, emmeans showed that our low performers in both groups recalled essentially the same number of critical lures (MSleep = 2.09, SDSleep = 1.63 vs.~Mwake = 2.09, SDWake = 1.74; \(\beta\) = 0.133, z = 0.021, p = .982). In other words, this exploratory analysis found no evidence that the effect of sleep on false recall was modulated by veridical memory, casting doubt over the robustness and reliability of Diekelmann et al.'s {[}\protect\hyperlink{ref-diekelmann2010a}{34}{]} findings.

@@

Across the Immediate and Delay groups, the number of intrusions and correct recall per list both had a main effect (ps \(\leq\) .001) such that they were positively correlated with lure recall. However, the effect of Test Time was not significant in either group (ps \textgreater{} .11). Finally, for the Correct recall per list \(\times\) Test Time interaction, it was significant in the Delay (z = -3.504, p \textless{} .001) but not the Immediate groups (z = -0.037, p = .971). To interpret the former, we used the R package, effects {[}\protect\hyperlink{ref-fox2022a}{\textbf{fox2022a?}}{]}, to visualise it (see Figure \ref{fig:delayfig}A) and plotted lure recall probability against veridical recall on a participant level (see Figure \ref{fig:delayfig}B).

>>>>

In this exploratory analysis, we followed Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{34}{]} by focusing on the Delay groups\footnote{Note that Diekelmann et al. (2010) did not have Immediate/Control groups.} and by performing a median split on our sample's adjusted veridical recall (Median = 5), classifying our participants as either high (\textgreater5; N = 117) or low (\(\leq\) 5; N = 123) performers. We followed the analysis approach of Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{34}{]}, but since the number of critical lures a participant produced is count data, we used a 2 (Group: Sleep vs.~Wake) \(\times\) 2 (Adjusted veridical recall: High vs.~Low Performers) Poisson regression, instead of an ANOVA. This analysis showed no main effect of Group (\(\beta\) = 0.052, z = 0.494, p = .621) but a main effect of Adjusted veridical recall (\(\beta\) = -0.438, z = -3.766, p \textless{} .001), such that high performers (M = 3.17, SD = 2.71) tended to falsely recall more critical lures than low performers (M = 2.09, SD = 1.68) ({[}\protect\hyperlink{ref-thapar2001a}{\textbf{thapar2001a?}},\protect\hyperlink{ref-toglia1999}{\textbf{toglia1999?}}{]}, but see {[}\protect\hyperlink{ref-roediger2001a}{21},\protect\hyperlink{ref-stadler1999a}{35},\protect\hyperlink{ref-cann2011a}{\textbf{cann2011a?}}{]}). Furthermore, importantly, the interaction between Group and Adjusted veridical recall was not significant (\(\beta\) = 0.049, z = 0.30, p = .763). Despite this, we followed Diekelmann et al. {[}\protect\hyperlink{ref-diekelmann2010a}{34}{]} by comparing low performers in the Sleep and Wake groups. Contrary to their findings, emmeans showed that our low performers in both groups recalled essentially the same number of critical lures (M\(_{Sleep}\) = 2.09, SD\(_{Sleep}\) = 1.63 vs.~M\(_{Wake}\) = 2.09, SD\(_{Wake}\) = 1.74; \(\beta\) = 0.133, z = 0.021, p = .982). In other words, this exploratory analysis found no evidence that the effect of sleep on false recall was modulated by veridical memory, casting doubt over the robustness and reliability of Diekelmann et al.'s {[}\protect\hyperlink{ref-diekelmann2010a}{34}{]} findings.

@@

Across the Immediate and Delay groups, the number of intrusions and correct recall per list both had a main effect (ps \(\leq\) .001) such that they were positively correlated with lure recall. However, the effect of Test Time was not significant in either group (ps \textgreater{} .11). Finally, for the Correct recall per list \(\times\) Test Time interaction, it was significant in the Delay (z = -3.471, p \textless.001) but not the Immediate groups (z = -0.037, p = .971). To interpret the former, we used the R package, effects {[}\protect\hyperlink{ref-fox2022a}{\textbf{fox2022a?}}{]}, to visualise it (see Figure \ref{fig:delayfig}A) and plotted lure recall probability against veridical recall on a participant level (see Figure \ref{fig:delayfig}B).
